Delayed effects of climate on vital rates leads to demographic divergence in Amazonian forest fragments

### 2021-05-17 doc/paper.Rmd
better now that we discussed the 2 paragraphs

institute:
  - UF_WEC: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430
  - Columbia: Department of Ecology, Evolution and Environmental Biology, Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA
  - UF_LAS: Center for Latin American Studies, University of Florida, Gainesville, Florida 32611-5530  
  - BDFFP: Biological Dynamics of Forest Fragments Project, INPA-PDBFF, CP 478, Manaus, Amazonas 69011-970 Brazil
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      linenumbers: yes
      number_sections: no
      fig_caption: yes
      reference_docx: "templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=templates/scholarly-metadata.lua
      - --lua-filter=templates/author-info-blocks.lua
      - --lua-filter=templates/pagebreak.lua
      - -Fpandoc-crossref
linkcolor: blue
bibliography: references.bib
csl: "templates/global-change-biology.csl" # Insert path for the bib-style
abstract: |
  Text of abstract. 34,281 year x individual measurements! 
keywords: |
  Amazon; delayed effects; demography; distributed lag non-linear models; drought; ENSO; habitat fragmentation; *Heliconia acuminata*; Heliconiaceae
highlights: |
  These are the highlights

@@

\pagebreak

@@

*Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot distributed throughout Central Amazonia [@kressDiversityDistributionHeliconia1990] and is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010]. While many *Heliconia* species grow in large patches in treefall gaps and other disturbed areas, others, such as *H. acuminata,* are found at lower densities in the darker and cooler forest understory [@rundelFunctionalTraitsBroadleaved2020]. These species produce fewer inflorescences and are pollinated by traplining rather than territorial hummingbirds [@stouffer1996; @brunaHeliconiaAcuminataReproductive2004].

@@

Each plot is subdivided into 50 quadrats ($10 \times 10$m) to simplify the annual surveys, during which we recorded the number of vegetative shoots each plant had, the height of each plant to the tallest leaf, and whether each plant was flowering (height and shoot number are correlated with leaf area, the probability of flowering, and rates of survivorship [@brunaHabitatFragmentationDemographic2002; @brunaEffectsForestFragmentation2002]).
In the current study we used the natural logarithm of the product of shoot number and height as a proxy for plant size.
Preliminary analysis showed that this was a better predictor of total leaf area than either shoot number or height alone (Table S2 <!--# update crossref manually -->).
Plants that were not found for three consecutive surveys were considered dead.
We also surveyed plots regularly during the rainy season to identify any that flowered after the survey.
For additional details on the plots, survey methods, and *H. acuminata* populations see @brunaHabitatFragmentationDemographic2002.

@@

Data on precipitation and potential evapotranspiration in our sites were obtained from a published gridded dataset (0.25º $\times$ 0.25º resolution) built using data from 3,625 ground-based weather stations across Brazil [@xavierDailyGriddedMeteorological2016].
We used these data to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010].
SPEI is a proxy for meteorological drought that integrates precipitation and evapotranspiration anomalies over a specified time scale.
Positive SPEI values for a given month indicate conditions wetter than the historical average for that month, while negative values of SPEI indicate droughts with intensity categorized as mild (0 to -1), moderate (-1 to -1.5), severe (-1.5 to -2), or extreme (\< -2) [@mckeeRelationshipDroughtFrequency1993].
SPEI can be calculated to represent different temporal scales of drought; we used 3-month SPEI because---given its shallow roots and rhizome---*H. acuminata* relies primarily on soil moisture rather than deeper water sources that can persist for longer timescales [@vicente-serranoMultiscalarDroughtIndex2010].
Note that 3-month SPEI is still monthly data---each month's SPEI value simply takes into account precipitation and evapotranspiration of the previous three months.
SPEI calculations were made using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017].
The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources, though the magnitude of drought can sometimes differ (Figure S2; Table S1 <!--# update crossrefs manually -->).

@@

To determine if plot characteristics influenced average vital rates we included a random effect of plot ID on the intercept; this was represented by $a_j$ in @eq:eqn1.
We determined the effects of SPEI on plant growth using plant size in year t+1 as a response variable.
This was modeled with a scaled t family error distribution because residuals were leptokurtic with a Gaussian error structure.
Because number of inflorescences was highly zero-inflated, we converted this to a binary response to model reproduction (i.e., 1 for ≥1 inflorescence, 0 for no inflorescences).
We modeled both reproduction and survival (i.e., from year t to year t+1) using a binomial family error distribution with a logit link function.
We modeled a potential cost of reproduction by including flowering in the previous year as covariate, $x_i$, in @eq:eqn1.

@@

Meteorological droughts in our focal region, as indicated by SPEI, are generally consistent with those reported in the literature (Table S1).
For example, the drought associated with the 1997 El Niño Southern Oscillation (ENSO) event was one of the most severe on record for the Amazon [@williamsonAmazonianTreeMortality2000]; correspondingly, 1997 has the lowest SPEI values in our timeseries (Figure \@ref(fig:eda)d). The 2005 dry season (June--October) was also reported as an exceptionally dry year, although this drought mostly affected the south western Amazon [@zengCausesImpacts20052008; @marengoDroughtAmazonia20052008].
Our SPEI data show the 2005 dry season to be a moderate drought (-1 \> SPEI \> -1.5).

@@

When summarizing across years and plots, the proportion of *Heliconia acuminata* that survived in CF and 1-ha was similarly high ($P_{surv} = 0.95$; Figure \@ref(fig:eda)b).
The proportion of surviving plants was lowest in the 2003--2004 transition year ($P_{surv} = 0.92$), which coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d) and was preceded by a drop in average plant size in the 2002--2003 transition year (Figure \@ref(fig:eda)a).
The lowest survival for 1-ha fragment plots ($P_{surv} = 0.93$) was for the 2005--2006 transition year, which encompassed a moderate drought in October 2005 and and wetter than average conditions (SPEI \> 0.5) in December 2005 and January 2006 (Figure \@ref(fig:eda)b,d).
Survival was size dependent in both continuous forest and 1-ha fragments ($p < 0.001$ for the effect of log-transformed plant size in year t on survival in year t+1 in both habitats).
While the survival probability of large plants approached 1 in both habitat types (Figure \@ref(fig:size-plot)a), the survival of the smallest plants was higher in 1-ha fragments.
However, the 95% confidence intervals for 1-ha fragments and continuous forest overlapped for all sizes.

Plants in continuous forest had an average of 2.9 shoots (± 1.8 SD) and were on average 40.6 cm tall (± 26.5 SD).
Plants in 1-ha fragments had on average \~`r format((1 - (2.5/2.9))*100, digits = 3)`% fewer shoots (2.5 ± 1.5 SD) and were \~`r format((1 - (36.2/40.6))*100, digits = 3)`% shorter (36.3 cm ± 24.1 SD).
Because the proxy for plant size used in our models was the product of these two metrics, plants in continuous forest were consequently substantially larger than those in forest fragments (150 ± 175 SD vs. 112 ± 141 SD, respectively).
<!--# Perhaps worth mentioning here that their distribution in different size classes is also different (skewed smaller) and likely why they are smaller on average). -->This disparity in plant size was most pronounced in the initial years of our surveys and then diminished over time (Figure \@ref(fig:eda)a), likely due to a combination of greater seedling recruitment in CF and plant mortality or shrinkage.
Mean plant size dropped dramatically in 2003 in both habitat types, corresponding with a severe drought during the February census (SPEI = -1.39) (Figure \@ref(fig:eda)d).
As with survival, size in year t was a significant predictor of size in year t+1 ($p < 0.001$ in both habitats).
While the effect was generally similar across size classes and habitat types, the impact of plant size on growth was greatest for mid-sized plants in continuous forest (Figure \@ref(fig:size-plot)b).

@@

While the overall proportion of all plants flowering was very low, it was nevertheless almost 40% higher in continuous forest than 1-ha fragments (`r summary_stats$flwr_mean[2]` ± `r summary_stats$flwr_sd[2]` vs. `r summary_stats$flwr_mean[1]` ± `r summary_stats$flwr_sd[1]`, respectively).
This disparity was largely due to the fact that flowering is size-dependent ($p < 0.001$ in both habitats), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 (log(size) \> 5 in Figure \@ref(fig:size-plot)c).
Even though flowering probability of large plants (log(size) \> 5) was greater in 1-ha fragments compared to continuous forest, there were proportionally far fewer plants above the reproductive threshold size in forest fragments (Figure \@ref(fig:size-plot)d).
Indeed, the most striking difference between habitat types coincided with a severe drought in 2003, when the percentage of flowering reproductive-sized plants was 28% in continuous forest vs. only 13.6% in 1-ha fragments (Figure \@ref(fig:eda)c).

@@

Our results indicate that some---but not all---results from short-term demographic studies can be extrapolated to longer times scales.
Decadal comparisons of *Heliconia acuminata* growth and survival largely mirrored those of prior experiments and surveys [@brunaHabitatFragmentationDemographic2002].
In contrast, conclusions regarding the effect of fragmentation on flowering differed substantially from those based on short-term surveys [@brunaHabitatFragmentationDemographic2002].
While in some years the probability a plant flowered was far lower in fragments than in continuous forest, in others it was comparable or even higher.
The probability that individual plants flowered continued to be size-dependent, with a similar size threshold below which plants were unlikely to produce inflorescences.
However, while in our initial three-year survey the relationship between plant size and probability of flowering was similar in both habitat types [@brunaHabitatFragmentationDemographic2002], when using the decadal dataset were found that flowering probability for the largest plants is actually higher in forest fragments (i.e., average fitted probability of flowering for plants in upper size quartile = 0.125 in CF vs. 0.094 in FF).

These size-dependent differences in flowering probability had only a modest effect on the previously observed disparities in flowering plant abundance---populations in fragments are smaller, and they have proportionately fewer of these large plants [@brunaHabitatFragmentationDemographic2002].
However, they could potentially lead to major habitat-specific differences in lifetime reproductive effort for these long-lived plants.
They translate into flowering once every 7.97 years in fragments vs. every 10.63 years in continuous forest (assuming stasis in size and average SPEI).
While some *Heliconia* species could compensate for these disparities with by selfing [@kressSelfincompatibilitySystemsCentral1983] or vegetative reproduction @rundelFunctionalTraitsBroadleaved2020], pollen transfer experiments indicate self-fertilization in *Heliconia acuminata* is rare and this species does not reproduce clonally (Bruna *unpubl. data*).
Our results underscore why caution is needed when using short-term studies to draw conclusions regarding the effects of fragmentation on long-lived organisms: seemingly small (i.e., non-significant) differences in vital rates, when compounded over multiple years, can lead to demographic divergence among populations.
[@gagnonGrowthUnderstoryHerb2011].

<!--# new paragraph below.  What do you think Emilio? -->

Fragmentation of forests diminishes their ability to buffer changes in microclimate due to edge effects [@didhamEdgeStructureDetermines1999; @ewersFragmentationImpairsMicroclimate2013].
This can result in changes in vegetation structure, and is a plausible mechanism for reduced growth and survival of *H. acuminata* in forest fragments in this study.
We show that the effects of variation in SPEI on survival and growth are more extreme in fragments compared to continuous forest.
Additionally, the direction of the SPEI effects appear to be seasonal---that is, extreme dry conditions in dry seasons and extreme wet conditions in wet seasons are most detrimental to growth and survival.
The ability of continuous forest habitat to buffer these extremes may be responsible for the less extreme effects of variation in SPEI in continuous forest.
More experimental work is needed to establish alterations in microclimate or impaired microclimate buffering in extreme weather events as a causal mechanism for reduced growth and survival of plant individuals in forest fragments.

@@

Precipitation extremes and other climatic factors are known to have immediate effects on the growth, survival, or reproduction of plants [@wrightSeasonalNinoLonger2006; @estebanOtherSideDroughts2021], including *Heliconia* [@stiles1975; @westerbandEarlyLifeConditions2017] and other tropical herbs [@wrightSeasonalDroughtSoil1992].
These effects can be complex or even contradictory---mild droughts can increase the growth rates of tropical trees and seedling survival, perhaps due to reductions in cloud cover and concomitant increases in solar radiation[@conditTropicalForestDynamics2004; @uriarteEnvironmentalHeterogeneityBiotic2018; @hueteAmazonRainforestsGreenup2006; @alfaro-sanchezGrowthReproductionRespond2017; @jonesAsynchronousAmazonForest2014], but in severe drought years growth can be extremely low and mortality can be sharply elevated [@connell2000; @engelbrecht2002; @edwards2006].
There is also evidence that the effects can persist for multiple years [@phillipsDroughtMortalityRelationships2010], such as a boom in drought-year fruit production followed by severe post-drought "famine" [@wright1999; @pau2013].

Despite this knowledge, models of plant population dynamics have rarely included the effects of environmental drivers.
This has largely been due to the challenge (both ecologically and statistically) of detecting any demographic responses to climatic extremes that are delayed for multiple growing seasons.
To address this challenge, researchers have begun to use a number of statistical methods that test for time lags in demographic responses without *a priori* assumptions about the influence of any particular climate window [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
Our expansion of this approach, which offers an unbiased way of identifying these delayed effects without overfitting (but see @pierreCriticalWindowsMethod2020 and @ogleQuantifyingEcologicalMemory2015 for alternative methods) yielded results consistent with this emerging literature---that the effects of precipitation extremes on the demography of *Heliconia acuminata* could be delayed for up to 3 growing seasons.

While it appears that delayed effects of climate on demographic vital rates may be ubiquitous [@eversLaggedDormantSeason2021], the extent to which they vary spatially or with habitat remains an open question.
Our results clearly indicate that they can, with habitat-specific differences in how environmental conditions influenced future vital rates.
For example, although extreme values of SPEI---both positive (unusually high precipitation) and negative (drought conditions)---led to declines in the probability of individual survival in both habitat types, the magnitude of these declines was far greater in forest fragments.
Similarly, the detrimental effects of extremes in SPEI on growth rates were also more pronounced in fragments.
In contrast, the effect of variation in SPEI on flowering probability was far stronger in continuous forest than fragments; these results should be interpreted with some caution, however, as the relatively low number of plants in fragments that are above the threshold-size for flowering could limit the power to detect delayed effects.

There are several, non-mutually-exclusive explanations for delayed effects of SPEI on demography.
The first is that the physiological processes underlying vital rates might be initiated long before they are demographically apparent [@eversLaggedDormantSeason2021], and hence be shaped by climatic events at any point in that physiological window.
For example, the flowering shoots of *Heliconia chartacea* begin to develop 6-10 months prior to the appearance of inflorescences.
Adverse conditions during the 6 months following initiation, rather than the months when inflorescences are starting expand, leads to the aborted production of flowering shoots [@crileyYearProductionHigh1994].
Interestingly, we observed the opposite effect---drought conditions increased the probability of flowering two years later.
While this could reflect bet-hedging in response to stress [@nihadRelationshipStressFlowering2018], this does not appear to be the case, as growth or survival do not appear to decrease following reproduction (see also [@horvitzDemographicCostReproduction1988]. In fact, flowering in one year is associated with increased reproduction and growth in the next.

@@

Third, it may be that the delayed demographic effects we observed are indirectly mediated by the effect of SPEI on other species rather than the direct effects on individual physiology [@eversLaggedDormantSeason2021].
For example, while droughts can kill some tropical trees immediately [@lauranceEffectsStrongDrought2001], others take three years or more [@crileyYearProductionHigh1994].
This will potentially lead to a boom in growth and flowering years after the drought, when the canopy gaps resulting from delayed leaf drop [@janssenDroughtEffectsLeaf2021] and treefalls allow for light penetration to the forest understory [@canham1990; @leitoldNinoDroughtIncreased2018].
Similar delayed changes in the local environment could also influence the foraging behavior of *H. acuminata*'s pollinators [@brunaHeliconiaAcuminataReproductive2004; @stouffer1996] and seed dispersers [@uriarteDisentanglingDriversReduced2011].
While more work is needed to explain why the (delayed) effects of SPEI on survival and growth are greater in forest fragments than interiors, it may be due to the proportionately greater amount of litterfall on edges [@vasconcelos2004] altering the abundance of pathogens or mycorrhizae.

@@

Finally, no analytical approach assessing the potential for demographic lags can compensate for a lack of long-term data [@tenhumbergTimelaggedEffectsWeather2018; @eversLaggedDormantSeason2021].
Unfortunately, long-term data monitoring the entire life-cycle of tropical taxa are rare, and those doing so in fragmented landscapes are virtually nonexistent [@brunaCompensatoryResponsesUnderstory2005].
Without investing in collecting such data, generalizations regarding the demographic consequences of climate change in these species rich and increasingly fragmented habitats will continue to prove elusive.

@@

```{r}
eda_cap <- "Timeseries of demographic parameters and drought occurrence. Mean plant size ± standard deviation (a) shows that plants were smaller in fragments (orange, solid) than continuous forest (blue, dashed). The proportion of plants surviving each transition year (b) shows that average survival is high and was lowest in the 2004 census. The proportion of reproductive sized plants (c) is on average low and fluctuates substantially year to year. The cutoff for reproductive size plants in panel (c) is defined as the upper 90th percentile size of flowering plants in all years. Monthly 3-month SPEI is shown (d) with gray lines representing values from different grid cells encompassing BDFFP and the dark line representing the site mean. Yellow, orange, dark orange, and red stripes show mild, moderate, severe, and extreme drought, respectively."
```

@@

-   @dowdInvestigatingSynergisticEffects2020 (unpublished thesis, could write MB and ask for update): precipitation and fragmentation influenced fruit/seed production (conditional on flowering) by *Heliconia* (one step past flowering, so might be worth citing to say that there are other ways reprod. can be affected

>>>>

institute:
  - UF_WEC: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430
  - Columbia: Department of Ecology, Evolution and Environmental Biology, Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA
  - UF_LAS: Center for Latin American Studies, University of Florida, Gainesville, Florida 32611-5530  
  - BDFFP: Biological Dynamics of Forest Fragments Project, INPA-PDBFF, CP 478, Manaus, Amazonas 69011-970 Brazil
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      linenumbers: yes
      number_sections: no
      fig_caption: yes
      reference_docx: "templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=templates/scholarly-metadata.lua
      - --lua-filter=templates/author-info-blocks.lua
      - --lua-filter=templates/pagebreak.lua
      - -Fpandoc-crossref
linkcolor: blue
bibliography: references.bib
csl: "templates/global-change-biology.csl" # Insert path for the bib-style
abstract: |
  Text of abstract. 34,281 year x individual measurments! 
keywords: |
  Amazon; delayed effects; demography; distributed lag non-linear models; drought; ENSO; habitat fragmentation; *Heliconia acuminata*; Heliconiaceae
highlights: |
  These are the highlights

@@

<!--# Insert page break -->

@@

*Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot distributed throughout Central Amazonia [@kressDiversityDistributionHeliconia1990] and is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010]. While many *Heliconia* species grow in large patches in trefall gaps and other disturbed areas, others, such as *H. acuminata,* are found at lower densities in the darker and cooler forest understory [@rundelFunctionalTraitsBroadleaved2020]. These species produce fewer infloresences and are pollinated by traplining rather than territorial hummingbirds [@stouffer1996; @brunaHeliconiaAcuminataReproductive2004].

@@

Each plot is subdivided into 50 quadrats ($10 \times 10$m) to simplify the annual surveys, during which we recorded the number of vegetative shoots each plant had, the height of each plant to the tallest leaf, and whether each plant was flowering (height and shoot number are correlated with leaf area, the probability of flowering, and rates of survivorship [@brunaHabitatFragmentationDemographic2002; @brunaEffectsForestFragmentation2002]).
Plants that were not found for three consecutive surveys were considered dead.
We also surveyed plots regularly during the rainy season to identify any that flowered after the survey.
For additional details on the location of plots, survey methods, and *H. acuminata* population structure see @brunaHabitatFragmentationDemographic2002.

@@

Data on precipitation and potential evapotranspiration in our sites were obtained from a published gridded dataset (0.25º $\times$ 0.25º resolution) built using data from 3,625 ground-based weather stations across Brazil [@xavierDailyGriddedMeteorological2016]. We used these data to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010].
SPEI is a proxy for meteorological drought that integrates precipitation and evapotranspiration anomalies over a specified time scale.
Positive SPEI values for a given month indicate conditions wetter than the historical average for that month, while negative values of SPEI indicate droughts with intensity categorized as mild (0 to -1), moderate (-1 to -1.5), severe (-1.5 to -2), or extreme (\< -2) [@mckeeRelationshipDroughtFrequency1993]. SPEI can be calculated to represent different temporal scales of drought; we used 3-month SPEI because---given its shallow roots and rhizome---*H. acuminata* relies primarily on soil moisture rather than deeper water sources that can persist for longer timescales [@vicente-serranoMultiscalarDroughtIndex2010].
Note that 3-month SPEI is still monthly data---each month's SPEI value simply takes into account precipitation and evapotranspiration of the previous three months.
SPEI calculations were made using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017].
The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources, though the magnitude of drought can sometimes differ (**Supplemental \_\_**).
<!--# add this as a "child document" or whatever -->

@@

To determine if plot characteristics influenced average vital rates we included a random effect of plot ID on the intercept; this was represented by $a_j$ in @eq:eqn1.
We determined the effects of SPEI on plant growth using plant size in year t+1 as a response variable.
This was modeled with a scaled t family error distribution and the natural log of height $\times$ shoot number as our proxy for plant size <!--# reference supplemental about plant size? -->.
A scaled t family was chosen because residuals were leptokurtic with a Gaussian error structure.
Because number of inflorescences was highly zero-inflated, we converted this to a binary response to model reproduction (i.e., 1 for ≥1 inflorescence, 0 for no inflorescences).
We modeled both reproduction and survival (i.e., from year t to year t+1) using a binomial family error distribution with a logit link function.
We modeled a potential cost of reproduction by including flowering in the previous year as covariate, $x_i$, in @eq:eqn1.

@@

Meteorological droughts in our focal region, as indicated by SPEI, are generally consistent with those reported in the literature.
For example, the drought associated with the 1997 El Niño Southern Oscillation (ENSO) event was one of the most severe on record for the Amazon [@williamsonAmazonianTreeMortality2000]; correspondingly, 1997 has the lowest SPEI values in our timeseries (Figure \@ref(fig:eda)d). The 2005 dry season (June--October) was also reported as an exceptionally dry year, although this drought mostly affected the south western Amazon [@zengCausesImpacts20052008; @marengoDroughtAmazonia20052008].
Our SPEI data show the 2005 dry season to be a moderate drought (-1 \> SPEI \> -1.5).

@@

When summarizing across years and plots, the proportion of *Heliconia acuminata* that survived in CF and 1-ha was similarly high ($P_{surv} = 0.95$; Figure \@ref(fig:eda)b).
The proportion of surviving plants was lowest in the 2003--2004 transition year ($P_{surv} = 0.92$), which coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d) and was preceded by a drop in average plant size in the 2002--2003 transition year (Figure \@ref(fig:eda)a).
The lowest survival for 1-ha fragment plots ($P_{surv} = 0.93$) was for the 2005--2006 transition year, which encompassed a moderate drought in October 2005 and and wetter than average conditions (SPEI \> 0.5) in December 2005 and January 2006 (Figure \@ref(fig:eda)b,d).
Survival was size dependent in both continuous forest and 1-ha fragments ($p < 0.001$ for the effect of log-transformed plant size in year t on suvival in year t+1 in both habitats).
While the survival probability of large plants approached 1 in both habitat types (Figure \@ref(fig:size-plot)a), the survival of the smallest plants was higher in 1-ha fragments.
However, the 95% confidence intervals for 1-ha fragments and continuous forest overlapped for all sizes.

Plants in continuous forest had an average of 2.9 shoots (± 1.8 SD) and were on average 40.6 cm tall (± 26.5 SD).
Plants in 1-ha fragments had on average \~`r format((1 - (2.5/2.9))*100, digits = 3)`% fewer shoots (2.5 ± 1.5 SD) and were \~`r format((1 - (36.2/40.6))*100, digits = 3)`% shorter (36.3 cm ± 24.1 SD).
Because the proxy for plant size used in our models was the product of these two metrics, plants in continuous forest were consequently substantially larger than those in forest fragments (150 ± 175 SD vs. 112 ± 141 SD, respectively).
<!--# Perhaps worth mentioning here that their distribution in different size classes is also different (skewed smaller) and likely why they are smaller on average). -->This disparity in plant size was most pronounced in the initial years of our surveys and then diminished over time (Figure \@ref(fig:eda)a).
Mean plant size dropped dramatically in 2003 in both habitat types, corresponding with a severe drought during the February census (SPEI = -1.39) (Figure \@ref(fig:eda)d).
As with survival, size in year t was a significant predictor of size in year t+1 ($p < 0.001$ in both habitats).
While the effect was generally similar across size classes and habitat types, the impact of plant size on growth was greatest for mid-sized plants in continuous forest (Figure \@ref(fig:size-plot)b).

@@

While the overall proportion of all plants flowering was very low, it was nevertheless almost 40% higher in continuous forest than 1-ha fragments (`r summary_stats$flwr_mean[2]` ± `r summary_stats$flwr_sd[2]` vs. `r summary_stats$flwr_mean[1]` ± `r summary_stats$flwr_sd[1]`, respectively).
This disparity was largely due to the fact that flowering is size-dependent ($p < 0.001$ in both habitats), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 (log(size) \> 5 in Figure \@ref(fig:size-plot)c) <!--# , which corresponds roughly to ____ shoots --This requires a supplemental figure, probably.  A plot of shoots vs. log(size) -->.
Even though flowering probability of large plants (log(size) \> 5) was greater in 1-ha fragments compared to continuous forest, there were proportionally far fewer plants above the reproductive threshold size in forest fragments (Figure \@ref(fig:size-plot)d).
Indeed, the most striking difference between habitat types coincided with a severe drought in 2003, when the percentage of flowering reproductive-sized plants was 28% in continuous forest vs. only 13.6% in 1-ha fragments (Figure \@ref(fig:eda)c).

@@

Many studies investigating the biological consequences of habitat fragmentation on plant growth, survival, and reproduction comprise short-term ([\<]{.ul}3 year) experiments and observations.
Our results underscore the difficulty in extrapolating long-term trends from such short-term studies, particularly when studying long-lived organisms or when the responses of interest can vary with size or age. For instance, one would have reached a very different conclusion regarding the effect of fragmentation on annual survival if the study windows were 1997-2002 (i.e., higher survival in continuous forest), 2002-2005 (i.e., higher survival in fragments), or 2004-2007 (i.e, no clear effect of fragmentation; \@ref(fig:eda)**X**). It is only when evaluating over longer time windows that it becomes apparent mortality is elevated in fragments relative fo continuous forest (**Fig 2**), and that the observed interannual variation is largely driven by dynamic patterns of recruitment [@brunaEffectsForestFragmentation2002] coupled with low mortality for plants beyond the smallest size classes [@bruna2003b].

Similarly, conclusions regarding the effects of fragmentation on flowering---which is also both rare and size-dependent [@brooks2019]---would also differ based on the year in which they were investigated. This could lead to erroneous extrapolations regarding the effects of fragmentation on reproductive mutualists or how population genetic structure [@uriarte2010; @uriarte2011; @cortes2013]. Conclusions based on short-term observations of temporally variable vital rates could lead to conservation and management practices that are ineffective or even counterproductive, especially when when failing to consider how the consequences of this variation might be modulated by organismal life history [@morris2008].

It is important to emphasize, however, that the overall effects of SPEI on survival and growth are more severe in fragments than continuous forest (**Fig X**).
Furthermore, the magnitude of plant responses to climatic extremes is also greater in habitat fragments ---extreme drought in dry seasons and extreme precipitation in during rainy seasons are most detrimental to growth and survival in fragments.
While intact forest and its canopy buffer populations from climatic extremes, populations in fragments - especially near edges with high contrast matrix - likely lack this protection [@didham1999;@ewers2013]. We suggest it is these climate extremes, rather than trends in average temperature, precipitation, or SPEI [@laurance2014], that that are the causal mechanism underlying reduced plant growth and survival in forest fragments.

@@

Climate anomalies are known to have immediate effects on the growth, survival, or reproduction of plants [@wrightSeasonalNinoLonger2006; @estebanOtherSideDroughts2021], including *Heliconia* [@stiles1975; @westerbandEarlyLifeConditions2017] and other tropical herbs [@wrightSeasonalDroughtSoil1992]. These effects can be complex or even contradictory---mild droughts can increase the growth rates of tropical trees and seedling survival, perhaps due to reductions in cloud cover and concomitant increases in solar radiation[@conditTropicalForestDynamics2004; @uriarteEnvironmentalHeterogeneityBiotic2018; @hueteAmazonRainforestsGreenup2006; @alfaro-sanchezGrowthReproductionRespond2017; @jonesAsynchronousAmazonForest2014], but in severe drought years growth can be extremely low and mortality can be sharply elevated [@connell2000; @engelbrecht2002; @edwards2006]. There is also evidence that the effects can persist for multiple years [@phillipsDroughtMortalityRelationships2010], such as a boom in drought-year fruit production followed by severe post-drought "famine" [@wright1999; @pau2013].

Despite these insights, models of plant population dynamics rarely include the effects of environmental drivers [but see @williams2015a; @tenhumberg2018; @molowny-horas2017). This has largely been due to the challenge (both ecologically and statistically) of detecting any demographic responses to climatic extremes that are delayed for multiple growing seasons. To address this, researchers have begun to use a number of statistical methods that test for time lags in demographic responses without *a priori* assumptions about the influence of any particular climate window [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
Our expansion of this approach, which offers an unbiased way of identifying these delayed effects without overfitting (but see @pierreCriticalWindowsMethod2020 and @ogleQuantifyingEcologicalMemory2015 for alternative methods) yielded results consistent with this emerging literature---that the effects of precipitation extremes on the demography of *Heliconia acuminata* could be delayed for up to 3 growing seasons.

While it appears that delayed effects of climate on demographic vital rates may be ubiquitous [@eversLaggedDormantSeason2021], the extent to which they vary spatially or with habitat remains an open question.
Our results clearly indicate that they can, with habitat-specific differences in how environmental conditions influenced future vital rates.
For example, extreme values of SPEI---both positive (unusually high precipitation) and negative (drought conditions)---led to declines in the probability of individual survival in both habitat types.
However, the magnitude of these declines was far greater in forest fragments.
Similarly, the detrimental effects of extremes in SPEI on growth rates were also more pronounced in fragments.
In contrast, variation in SPEI had far stronger effects on the probability of flowering in continuous forest than fragments.
These results should be interpreted with some caution, however, as the relatively low number of plants in fragments that are above the threshold-size for flowering could limit the power to detect delayed effects.

There are several, non-mutually-exclusive explanations for delayed effects of SPEI on demography.
The first is that the physiological processes underlying vital rates might be initiated long before they are demographically apparent [@eversLaggedDormantSeason2021], and hence be shaped by climatic events at any point in that physiological window. For example, the flowering shoots of *Heliconia chartacea* begin to develop 6-10 months prior to the appearance of infloresences. Adverse conditions during the 6 months following initiation, rather than the months when infloresences are starting expand, leads to the aborted production of flowering shoots [@crileyYearProductionHigh1994]. Interestingly, we observed the opposite effect---drought conditions increased the probability of flowering two years later. While this could reflect bet-hedging in response to stress [@nihadRelationshipStressFlowering2018], this does not appear to be the case, as growth or survival do not appear to decrease following reproduction (see also [@horvitzDemographicCostReproduction1988]. In fact, flowering in one year is associated with increased reproduction and growth in the next.

@@

Third, it may be that the delayed demographic effects we observed are indirectly mediated by the effect of SPEI on other species rather than the direct effects on individual physiology [@eversLaggedDormantSeason2021]. For example, topical trees may not die until three or more years after a drought [@crileyYearProductionHigh1994]. When they finally do, the resulting leaf drop [@janssenDroughtEffectsLeaf2021] and treefalls allow for light penetration to the forest understory [@canham1990; @leitoldNinoDroughtIncreased2018], triggering a boom in the growth and flowering of understory plants [@bruna2005a]. Similar delayed changes in the local environment could also influence the foraging behavior of a plant's pollinators [@brunaHeliconiaAcuminataReproductive2004; @stouffer1996], seed dispersers [@uriarteDisentanglingDriversReduced2011], or herbivores [@scotta]. While more work is needed to explain why the (delayed) effects of SPEI on *H. acuminata* survival and growth are greater in fragments than forest interiors, one hypothesis motivated by recent intriguing results from other systems **(CITATIONS)** is that the greater litterfall on edges [@vasconcelos2004] may be altering the abundance of pathogens or mycorrhizae.

@@

Finally, no analytical approach assessing the potential for demographic lags can compensate for a lack of long-term data [@tenhumbergTimelaggedEffectsWeather2018; @eversLaggedDormantSeason2021]. Unfortunately, long-term data monitoring the entire life-cycle of tropical taxa are rare, and those doing so in fragmented landscapes are virtually nonexistent [@brunaCompensatoryResponsesUnderstory2005].
Without investing in collecting such data, generalizations regarding the demographic consequences of climate change in these species rich and increasingly fragmented habitats will continue to prove elusive.
More generally, however, researchers need to consider how delayed responses to climate could influence the interpretation of data in studies where the organisms lifespan exceeds the study's duration.

@@

```{r}
eda_cap <- "Timeseries of demographic parameters and drought ocurrence. Mean plant size ± standard deviation (a) shows that plants were smaller in fragments (orange, solid) than continuous forest (blue, dashed). The proportion of plants surviving each transition year (b) shows that average survival is high and was lowest in the 2004 census. The proportion of reproductive sized plants (c) is on average low and fluctuates substantially year to year. The cutoff for reproductive size plants in panel (c) is defined as the upper 90th percentile size of flowering plants in all years. Monthly 3-month SPEI is shown (d) with grey lines representing values from different grid cells encompassing BDFFP and the dark line representing the site mean. Yellow, orange, dark orange, and red stripes show mild, moderate, severe, and extreme drought, respectively."
```

@@

-   @dowdInvestigatingSynergisticEffects2020 (unpublished thesis, could write MB and ask for update): precipitation and fragmentations influenced fruit/seed production (conditional on flowering) by *Heliconia* (one step past flowering, so might be worth citing to say that there are other ways reprod. can be affected
